---
output:
  pdf_document: default
  html_document: default
---
#' Author: Mohamed Yehia Mokhtar Salama
#' Date: October 6th, 2024
#' Purpose: OKCupid Case I
#' 

#### Importing Necessary Libraries
```{r}
#install.packages("ggplot2")
library(ggplot2)
library(dplyr)
```

#### Setting My WD
```{r}
setwd("/cloud/project")
```

#### Seeing All Files in WD
```{r}
dir()
```

#### Importing All Data
```{r}
profiles <- read.csv('profiles.csv')
latlon <- read.csv('LatLon.csv')
shared_census <- read.csv("sharedCensus2010Vars.csv")
address <- read.csv("addr.csv")
```

#### Making Sure Data Is Loaded and Displaying Head
```{r}
head(profiles)
head(latlon)
#head(shared_census)
head(address)
```
#### Summary statistics for all datasets
```{r}
summary(profiles)
summary(latlon)
#summary(shared_census)
summary(address)
```
#### Checking Datatypes
```{r}
str(profiles)
str(latlon)
#str(shared_census)
str(address)
```
# Missing Values Check
```{r}
colSums(is.na(profiles))
colSums(is.na(latlon))
#colSums(is.na(shared_census))
colSums(is.na(address))
```
#### Converting Necessary Datatypes
```{r}
# Convert character columns to factors for 'profiles' dataset
profiles$body_type <- as.factor(profiles$body_type)
profiles$diet <- as.factor(profiles$diet)
profiles$drinks <- as.factor(profiles$drinks)
profiles$drugs <- as.factor(profiles$drugs)
profiles$education <- as.factor(profiles$education)
profiles$ethnicity <- as.factor(profiles$ethnicity)
profiles$job <- as.factor(profiles$job)
profiles$offspring <- as.factor(profiles$offspring)
profiles$orientation <- as.factor(profiles$orientation)
profiles$pets <- as.factor(profiles$pets)
profiles$religion <- as.factor(profiles$religion)
profiles$sex <- as.factor(profiles$sex)
profiles$sign <- as.factor(profiles$sign)
profiles$smokes <- as.factor(profiles$smokes)
profiles$speaks <- as.factor(profiles$speaks)
profiles$status <- as.factor(profiles$status)

# Convert address columns to factors
address$location <- as.factor(address$location)
address$city <- as.factor(address$city)
address$county <- as.factor(address$county)
address$state <- as.factor(address$state)

# Convert 'last_online' to a datetime object
profiles$last_online <- as.POSIXct(profiles$last_online, format = "%Y-%m-%d %H:%M:%S")

# Checking the structure of profiles after conversion
str(profiles)
```

#### Some Exploration: Distribution Plots
```{r}
# For Profiles Dataset
numeric_vars_profiles <- sapply(profiles, is.numeric)
profiles_numeric <- profiles[, numeric_vars_profiles]
par(mfrow = c(2, 2)) # Set up plotting area for multiple plots
for (var in names(profiles_numeric)) {
  hist(profiles_numeric[[var]], main = paste("Distribution of", var), xlab = var, col = "blue")
}
```

#### Some Exploration: Bar Plots for Categorical Variables
```{r}
# Profiles Dataset
categorical_vars_profiles <- sapply(profiles, is.factor)
profiles_categorical <- profiles[, categorical_vars_profiles]
par(mfrow = c(2, 2))
for (var in names(profiles_categorical)) {
  barplot(table(profiles_categorical[[var]]), main = paste("Bar Plot of", var), xlab = var, col = "blue")
}
```

#### Handling Missing 'income' Values
```{r}
# Checking how many missing values exist in 'income'
sum(is.na(profiles$income))

# Imputeing missing 'income' values with the mean of non-missing values
profiles$income[is.na(profiles$income)] <- mean(profiles$income, na.rm = TRUE)
```

#### Feature Engineering for Status and Education
```{r}
# Feature engineering a new column 'statEDU' by combining 'status' and 'education'
profiles$statEDU <- paste(profiles$status, profiles$education, sep = '_')

# Displaying the distribution of the new feature
table(profiles$statEDU)
```

#### Joining All Data
```{r}
# Enrich 'profiles' data by joining with 'LatLon.csv' on the 'location' column
moreData <- left_join(profiles, latlon, by = 'location')

# Join with sharedCensus2010Vars dataset
#moreData_enriched <- left_join(moreData, shared_census, by = 'location')

# Join with address dataset
moreData_enriched <- left_join(moreData, address, by = 'location')

head(moreData_enriched)
```

#### Handling Missing Data Using 'complete.cases()' and Dropping Columns with 50% or more missing values
```{r}
# Calculate the percentage of missing values for each column
missing_percentage <- colSums(is.na(moreData_enriched)) / nrow(moreData_enriched) * 100

# Convert the missing percentage data to a data frame for visualization
missing_df <- data.frame(variable = names(missing_percentage), 
                         missing_percent = missing_percentage)

# Plot the missing percentages
ggplot(missing_df, aes(x = reorder(variable, -missing_percent), y = missing_percent)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +  
  labs(title = "Percentage of Missing Data by Variable",
       x = "Variables", 
       y = "Percentage of Missing Data") +
  theme_minimal()

# Dropping Columns with High Percentage of Missing Values
# Deciding to drop columns with more than 50% missing values
threshold <- 50 
columns_to_drop <- names(missing_percentage[missing_percentage > threshold])

# Dropping these columns from the dataset
moreData_cleaned <- moreData_enriched[, !(names(moreData_enriched) %in% columns_to_drop)]

# Check the new dataset
str(moreData_cleaned)

#### Handling Missing Data Using 'complete.cases()'
# Create a new dataset that includes only complete cases (rows with no missing values)
completeMoreData <- moreData_cleaned[complete.cases(moreData_cleaned),]

# Check the number of rows in the complete dataset
nrow(completeMoreData)

# Check the head of the new data
head(completeMoreData)

str(completeMoreData)
```
```{r}
# Age vs Drinks: Density plot to show how drinking habits vary across different age groups
ggplot(completeMoreData, aes(x = age, fill = drinks)) +
  geom_density(alpha = 0.6) +  
  labs(title = "Age Distribution by Drinking Habits", x = "Age", fill = "Drinks") +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "#5DC0F8", color = NA),  
    plot.background = element_rect(fill = "#5DC0F8", color = NA),   
    panel.grid.major = element_blank(),  
    panel.grid.minor = element_blank(),  
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),  
    legend.position = "right",  
    axis.title = element_text(size = 14, face = "bold"),  
    axis.text = element_text(size = 12, face = "bold")    
  ) +
  scale_fill_manual(values = c(
    "desperately" = "#FF6347",  
    "not at all" = "#4682B4",   
    "often" = "#32CD32",        
    "rarely" = "#FFD700",       
    "socially" = "#8A2BE2",     
    "very often" = "#FF4500"    
  ))
```

```{r}
# Body Type vs Diet: Stacked bar chart to show relationship between body type and diet preferences
ggplot(completeMoreData, aes(x = body_type, fill = diet)) +
  geom_bar(position = "fill") + 
  labs(title = "Diet Preferences by Body Type", x = "Body Type", fill = "Diet Preferences") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Religion vs Drinks: Stacked bar chart to show relationship between religion and drinking habits
ggplot(completeMoreData, aes(x = religion, fill = drinks)) +
  geom_bar(position = "fill") +  # Stacked bar chart for religion vs drinking
  labs(title = "Drinking Habits by Religion", x = "Religion", fill = "Drinks") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
# Smoking vs Diet: Stacked bar chart to show the relationship between smoking habits and diet preferences
ggplot(completeMoreData, aes(x = diet, fill = smokes)) +
  geom_bar(position = "fill") + 
  labs(title = "Smoking Habits by Diet Preferences", x = "Diet", fill = "Smoking Habits") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# Pets vs Relationship Status: Stacked bar chart to show the relationship between pet preferences and relationship status
ggplot(completeMoreData, aes(x = pets, fill = status)) +
  geom_bar(position = "fill") +  
  labs(title = "Relationship Status by Pet Preferences", x = "Pets", fill = "Relationship Status") + 
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "#5DC0F8", color = NA),  
    plot.background = element_rect(fill = "#5DC0F8", color = NA),   
    panel.grid.major = element_blank(),  
    panel.grid.minor = element_blank(),  
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),  
    legend.position = "right",  
    axis.title = element_text(size = 14, face = "bold"),  
    axis.text = element_text(size = 12, face = "bold"),   
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)  
  ) +
  scale_fill_manual(values = c(
    "available" = "#FF6347",    
    "married" = "#4682B4",      
    "seeing someone" = "#32CD32",
    "single" = "#FFD700",       
    "unknown" = "#8A2BE2"       
  ))
```

```{r}
# Calculating the total number of entries
total_entries <- nrow(completeMoreData)

# Calculating the count for each pet category
pet_counts <- table(completeMoreData$pets)

# Calculating the percentage for each category
pet_percentages <- (pet_counts / total_entries) * 100

pet_percentages
```
```{r}
# Summarizing the percentages for each category in the "pets" column
pet_counts <- completeMoreData %>%
  group_by(pets) %>%
  summarise(count = n()) %>%
  mutate(percentage = (count / sum(count)) * 100)

# Creating the pie chart with conditional labels
ggplot(pet_counts, aes(x = "", y = percentage, fill = pets)) +
  geom_bar(stat = "identity", width = 1, color = "white") + 
  coord_polar("y") +  
  labs(title = "Percentage Distribution of Pet Preferences", fill = "Pets") +
  theme_minimal() +  
  theme(
    panel.background = element_rect(fill = "#5DC0F8", color = NA),  
    plot.background = element_rect(fill = "#5DC0F8", color = NA),   
    panel.grid.major = element_blank(),  
    panel.grid.minor = element_blank(),  
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"), 
    axis.title.x = element_blank(),  
    axis.title.y = element_blank(),  
    axis.text = element_blank(),  
    axis.ticks = element_blank()  
  ) +
  scale_fill_manual(values = c(
    "dislikes cats" = "#FF6347",  
    "dislikes dogs" = "#4682B4",  
    "dislikes dogs and dislikes cats" = "#32CD32",  
    "dislikes dogs and has cats" = "#FFD700",  
    "dislikes dogs and likes cats" = "#8A2BE2",  
    "has cats" = "#D2691E",  
    "has dogs" = "#FF4500",  
    "has dogs and dislikes cats" = "#1E90FF",  
    "has dogs and has cats" = "#7FFF00",  
    "has dogs and likes cats" = "#9370DB",  
    "likes cats" = "#FF69B4",  
    "likes dogs" = "#B22222",  
    "likes dogs and dislikes cats" = "#3CB371",  
    "likes dogs and has cats" = "#DA70D6",  
    "likes dogs and likes cats" = "#FFD700"  
  )) +
  geom_text(aes(label = ifelse(pets == "dislikes dogs and dislikes cats" | percentage >= 5, 
                               paste0(round(percentage, 1), "%"), "")), 
            position = position_stack(vjust = 0.5), 
            size = 5,              
            fontface = "bold",     
            color = "black")       
```

```{r}
ggplot(completeMoreData, aes(x = status, y = income, fill = status)) +
  geom_boxplot(outlier.colour = "black", outlier.shape = 16, outlier.size = 2, notch = FALSE) +
  labs(title = "Income Distribution by Relationship Status", x = "Relationship Status", y = "Income") +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "#5DC0F8", color = NA),  
    plot.background = element_rect(fill = "#5DC0F8", color = NA),   
    panel.grid.major = element_blank(),  
    panel.grid.minor = element_blank(),  
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),  
    legend.position = "none",  
    axis.title = element_text(size = 14, face = "bold"),  
    axis.text = element_text(size = 12, face = "bold")    
  ) +
  scale_fill_manual(values = c(
    "available" = "#FF6347",    
    "married" = "#4682B4",      
    "seeing someone" = "#32CD32",
    "single" = "#FFD700",       
    "unknown" = "#8A2BE2"       
  ))
```

```{r}
# Grouped boxplot for Income vs Job grouped by Relationship Status
ggplot(completeMoreData, aes(x = job, y = income, fill = status)) +
  geom_boxplot(outlier.colour = "black", outlier.shape = 16, outlier.size = 2, notch = FALSE) +
  labs(title = "Income Distribution by Job and Relationship Status", x = "Job", y = "Income") +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "#5DC0F8", color = NA),  
    plot.background = element_rect(fill = "#5DC0F8", color = NA),   
    panel.grid.major = element_blank(),  
    panel.grid.minor = element_blank(),  
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),  
    axis.title = element_text(size = 14, face = "bold"),  
    axis.text = element_text(size = 12, face = "bold"),   
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)  
  ) +
  scale_fill_manual(values = c(
    "available" = "#FF6347",    
    "married" = "#4682B4",      
    "seeing someone" = "#32CD32",
    "single" = "#FFD700",       
    "unknown" = "#8A2BE2"       
  ))
```

```{r}
# Grouped boxplot for Income vs Orientation colored by Status
ggplot(completeMoreData, aes(x = orientation, y = income, fill = status)) + 
  geom_boxplot(outlier.colour = "black", outlier.shape = 16, outlier.size = 2, notch = FALSE) + 
  labs(title = "Income Distribution by Orientation and Relationship Status", 
       x = "Orientation", 
       y = "Income", 
       fill = "Relationship Status") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  scale_fill_manual(values = c(
    "available" = "#FF6347",    
    "married" = "#4682B4",      
    "seeing someone" = "#32CD32",
    "single" = "#FFD700",       
    "unknown" = "#8A2BE2"       
  ))
```

```{r}
# Plot for Age Distribution by Gender and Relationship Status
ggplot(completeMoreData, aes(x = age, fill = status)) +
  geom_density(alpha = 0.6) + 
  facet_wrap(~sex) +  
  labs(title = "Age Distribution by Gender and Relationship Status", x = "Age", fill = "Relationship Status") +
  theme_minimal() + 
  theme(
    panel.background = element_rect(fill = "#5DC0F8", color = NA),  
    plot.background = element_rect(fill = "#5DC0F8", color = NA),   
    panel.grid.major = element_blank(),  
    panel.grid.minor = element_blank(),  
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),  
    legend.position = "right",  
    axis.title = element_text(size = 14, face = "bold"),  
    axis.text = element_text(size = 12, face = "bold")    
  ) +
  scale_fill_manual(values = c(
    "available" = "#FF6347",    
    "married" = "#4682B4",      
    "seeing someone" = "#32CD32",
    "single" = "#FFD700",       
    "unknown" = "#8A2BE2"       
  ))
```

```{r}
# Getting the total number of users
total_users <- nrow(completeMoreData)

# Getting the top 10 cities with the most users and calculating the percentage
top_cities <- completeMoreData %>%
  group_by(city) %>%
  summarise(user_count = n()) %>%
  mutate(percentage = (user_count / total_users) * 100) %>%
  arrange(desc(user_count)) %>%
  top_n(10, user_count)

# Plot Top 10 Cities by User Count with Percentage Labels
ggplot(top_cities, aes(x = reorder(city, -user_count), y = user_count)) +
  geom_bar(stat = "identity", fill = "#FFD700") +  
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            vjust = -0.5, size = 5, color = "black", fontface = "bold") +  
  labs(title = "Top 10 Cities with Most Users", x = "City", y = "Count of Users") +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "#5DC0F8", color = NA),  
    plot.background = element_rect(fill = "#5DC0F8", color = NA),   
    panel.grid.major = element_blank(),  
    panel.grid.minor = element_blank(),  
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),  
    legend.position = "none",  
    axis.title = element_text(size = 14, face = "bold"),  
    axis.text = element_text(size = 12, face = "bold"),   
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)  
  )
```

```{r}
# Calculating the percentages for each education status
total_users <- nrow(completeMoreData)

# Grouping by the education status and calculateing the count and percentage
top_edu_statuses <- completeMoreData %>%
  group_by(statEDU) %>%
  summarise(user_count = n()) %>%
  mutate(percentage = (user_count / total_users) * 100) %>%
  arrange(desc(user_count)) %>%
  top_n(10, user_count)

ggplot(top_edu_statuses, aes(x = reorder(statEDU, -percentage), y = percentage)) +
  geom_bar(stat = "identity", fill = "#FFD700") +  
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            vjust = -0.5, size = 5, color = "black", fontface = "bold") +  
  labs(title = "Top 10 Education Statuses", x = "Education Status", y = "Percentage") +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "#5DC0F8", color = NA),  
    plot.background = element_rect(fill = "#5DC0F8", color = NA),   
    panel.grid.major = element_blank(),  
    panel.grid.minor = element_blank(),  
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),  
    legend.position = "none",  
    axis.title = element_text(size = 14, face = "bold"),  
    axis.text = element_text(size = 12, face = "bold"),   
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)  
  )

```

```{r}
# Filter the top 10 cities with the most users
top_cities <- completeMoreData %>%
  group_by(city) %>%
  summarise(user_count = n()) %>%
  arrange(desc(user_count)) %>%
  top_n(10, user_count)

# Filter data to only include users from the top 10 cities
filtered_data <- completeMoreData %>% filter(city %in% top_cities$city)

# Boxplot for Age distribution by Top 10 Cities
ggplot(filtered_data, aes(x = reorder(city, age, FUN = median), y = age)) + 
  geom_boxplot(fill = "#FFD700", color = "black", outlier.colour = "red", outlier.size = 2) +
  labs(title = "Age Distribution by Top 10 Cities", x = "City", y = "Age") +
  coord_flip() +  # Flip the axes
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "#5DC0F8", color = NA),  
    plot.background = element_rect(fill = "#5DC0F8", color = NA),   
    panel.grid.major = element_blank(),  
    panel.grid.minor = element_blank(),  
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),  
    axis.title = element_text(size = 14, face = "bold"),  
    axis.text = element_text(size = 12, face = "bold")    
  )
```